---
title: "Solid Waste Facilities Near Seed Manufacturer"
author: "Barbara Monaco"
date: "8/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rgdal)
library(tidyverse)
library(leaflet)
library(fontawesome)
library(sf)
library(units)
```
Exploration of existing solid waste facility data to identidy the most likely (closest) landfills, transfer station or WTE/RDF facilities that the below companies would be using.

```{r, include = FALSE}
fgdb <- "R:/mpca/util_solid_waste.gdb"

fc_list <- ogrListLayers(fgdb)

# Read the feature class
fc <- readOGR(dsn=fgdb,layer="solid_waste") %>% 
  st_as_sf()

#list of AIs of WTE/RDF facilities are also included in the 
#filter statement, then to get one location per facility, we use the
#group by and summarise statements to get distinct facilities
transfer_msw_ld <- fc %>% 
  filter(si_type_desc %in% c("Transfer Station",
                             "Transfer Area",
                             "MSW Land Disposal")|ai_id %in% 
           c(1143,726,4218,1156,4386,736,1609,40520)) %>% 
  group_by(ai_id, ai_name, permit, address1, city, state, zip) %>% 
  summarise(longitude = max(longitude), latitude = max(latitude))
```


```{r}
#Locations were determined using Google maps which uses the webmercator
#projection.
projcrs <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

treated_seed_locs <- read.csv("https://raw.githubusercontent.com/MPCA-data/tidytuesdays/master/show-n-tell/maps/Treated%20Seed%20Locations%208-12-21.csv")
treated_seed_sf <- st_as_sf(x = treated_seed_locs,
                              coords = c("Long", "Lat"),
                              crs = projcrs,
                            remove = FALSE)

#We need both datasets to be in the same projection so we transform
#the treated seed locations to UTM to match our facility data
treated_seed_sf <- st_transform(treated_seed_sf, crs = 26915)
#Calculates distances in miles. Each row is a seed mfg and each
#column is a SW facility
distances <- st_distance(treated_seed_sf, transfer_msw_ld) %>% 
  set_units("miles")

t_distances <- t(distances) %>% as.data.frame() %>% round(2)
names(t_distances) <- paste(treated_seed_locs$Entity.Name, "Distance")
combined_data <- bind_cols(transfer_msw_ld, t_distances)
```

```{r}
treated_seed_locs %>% 
  dplyr::select(Name = Entity.Name, Address = Address1, City, County, 
         Zip = Postal.Code, Comments) %>% 
  knitr::kable()
```
# Map of Facilities

Use the ruler icon to draw and measure the distances between one point and another, you can zoom and click on each marker to bring up a pop-up for each facility/location.
```{r}
trash_icons <- awesomeIcons(
  icon = 'trash',
  iconColor = 'black',
  library = 'fa',
  markerColor = 'green'
)

seed <- awesomeIcons(text = fa('seedling', height = '22px'),                          
                     iconColor   = 'green', 
                     library     = 'fa',
                     markerColor = 'white')

leaflet(data = transfer_msw_ld) %>%
  addTiles() %>% 
  addAwesomeMarkers(~longitude, ~latitude, icon = trash_icons, 
                    popup = ~paste0(permit, "<br>" ,ai_name, 
                                    "<br>", address1, "<br>",
                                    city, ", MN ", zip)) %>% 
  addAwesomeMarkers(data = treated_seed_locs, ~Long, ~Lat, 
                    icon = seed, popup = ~paste0(Entity.Name, "<br>",
                                                 Address1, "<br>",
                                                 City, ", MN ", Postal.Code)) %>% 
  addMeasure()

```

Below are the top 5 facilities closest to each of the locations with their address and distances (in miles) reported

## AgReliant Genetics LLC
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("AgReli")) %>% 
  slice_min(order_by = `AgReliant Genetics LLC Distance`, n = 5) %>% 
  knitr::kable()
```

## Centra Sota Cooperative
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("Centra S")) %>% 
  slice_min(order_by = `Centra Sota Cooperative Distance`, n = 5) %>% 
  knitr::kable()
```

## Dahlco Seeds Inc
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("Dahlco")) %>% 
  slice_min(order_by = `Dahlco Seeds Inc Distance`, n = 5) %>% 
  knitr::kable()
```

## Dekalb
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("Dekalb")) %>% 
  slice_min(order_by = `Dekalb Distance`, n = 5) %>% 
  knitr::kable()
```

## Dupont Pioneer
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("DuPont")) %>% 
  slice_min(order_by = `DuPont Pioneer Distance`, n = 5) %>% 
  knitr::kable()
```

## Monsanto Co
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("Monsanto")) %>% 
  slice_min(order_by = `Monsanto Co Distance`, n = 5) %>% 
  knitr::kable()
```

## Remington Seeds
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("Reming")) %>% 
  slice_min(order_by = `Remington Seeds Distance`, n = 5) %>% 
  knitr::kable()
```

## Syngenta Seeds, LLC
```{r}
combined_data %>% 
  as.data.frame() %>% 
  dplyr::select(Name = ai_name, Permit = permit, Address = address1, 
                City = city, State = state, Zip = zip, 
                starts_with("Syngenta")) %>% 
  slice_min(order_by = `Syngenta Seeds, LLC Distance`, n = 5) %>% 
  knitr::kable()
```



