---
title: "Zotero citation primer"
author: "Me^1^, Someone else^2,\\*^"
date: "^1^I UCS, another place"
output:
  word_document:
    keep_md: true

bibliography: My library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

I am writing and may want to cite something, [@amiriWorkingEnvironmentalJustice2019] And, then sometimes I just want the citation to show up in the text, [@alvesEPAAuthorityUse2012] And then I will knit this to show you what it looks like in the R Markdown document. 

As I am typing and writing my paper, I come upon something I want to cite  [@arquetteHolisticRiskbasedEnvironmental2002].

These are the youtube videos and primers I read to learn how to do this:

This one helps you install Zotero and Better-bibtext:
https://dewey.dunnington.ca/post/2020/getting-started-zotero-better-bibtex-rmarkdown/

This got me over the hump in learning how to set up Zotero,
https://www.youtube.com/watch?v=Nm5BsfMfO9M

Here is the Zotero webpage. You will have to download Zotero and create a Zotero account online. That way you can share your bibs with other authors.
https://www.zotero.org/

And better-bibtex download and installation instructions are here,
https://retorque.re/zotero-better-bibtex/

You will need the R Studio add-in, citr. Due to COVID related timing issues, citr is not updated on CRAN, but can be downloaded using devtools. 

That documentation is here: 
https://github.com/crsh/citr

When you are knitting and re-knitting the Word Document, you have to close it or R Studio will throw an error.

```{r packages}

#devtools::install_github("crsh/citr")

library(citr)

```

