---
title: 'Tutorial: tidycensus'
author: "Amanda Gorton, MPCA"
date: "2022-09-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

### Why tidycensus?  

- Access to Decennial Census, American Community Survey (ACS), Population Estimates Program, Public Use Microdata Series APIs  
- Mapping is way easier because census geometries are automatically downloaded  
- States and counties can be requested by name rather than FIPs codes   

Note that there is **excellent** documentation for tidycensus, including a FREE book by Kyle Walker, who created tidycensus!  
https://walker-data.com/census-r/index.html  


### Getting started 

Before we do anything, let's load our packages. Remember that if you don't have these installed first, you will get an error when you run the code below. 

#### Load packages

```{r}
library(tidycensus)
library(dplyr)
library(ggplot2)
library(keyring)
library(tidyr)
```

#### Get your API key + load

In order to access census data, you will need to request your own census API key. This can be done [here]:  
(https://api.census.gov/data/key_signup.html)   

```{r, eval = FALSE}
census_api_key("YOUR API KEY HERE")
```

Alternatively, if you use keychain (and better for sharing code), you can hide your API key and not have to paste/enter it each time   
```{r, eval = FALSE}
# Store your individual API key
key_set_with_value(service = "census api", 
                   password = "YOUR API KEY HERE") 

```

```{r}
census_api_key(key_get("census api")) # If you include, install = TRUE, it will install your API key for use in future sessions
```


### Basics of tidycensus

The two major functions used in tidycensus to pull in census data (2000,2010,2020) and ACS data (5 year, 1 year) are as follows:  

`get_decennial()`  
`get_acs()`  

Within these functions, you can specify the geography (state, county, tract, block, etc), the census or ACS variable of interest, and the year. There are other functions available as well - see documentation!    

A full list of available geographies is provided here, although note that not all geographies are available for all variables:  

https://walker-data.com/tidycensus/articles/basic-usage.html  

Here are some examples:

```{r}
# Median age by state for the 2010 Census 

state_median_age <- get_decennial(geography = "state", 
                                  variables = "P013001", 
                                  year = 2010)
head(state_median_age)
```

```{r}
# Income estimates by tract, for Hennepin county only. ACS 2016-2020

income_hennepin_tracts <- get_acs(state = "MN",
                                  county = "Hennepin",
                                  geography = "tract",
                                  variables = "B19013_001",
                                  year = 2020)

head(income_hennepin_tracts)

```

#### Exploring Census and ACS variables 

As you may have noticed, the two examples above involved knowing the variable ID of interest. This can be tricky since there are hundreds of these!  

This is another place where tidycensus is very handy. The function `load_variables()` allows us to view and search all the variables in R Studio.  

You can also use something like Census Reporter (https://censusreporter.org/) to browse tables of interest

You have to specify the year, and then the dataset of interest.More details here:  
https://walker-data.com/tidycensus/reference/load_variables.html   

Let's load some from the 2010 census and browse:
```{r}
census.2010 <- load_variables(2010, "sf2", cache = TRUE)

```

Load some ACS data and browse:
```{r}
acs.2020 <- load_variables(2020, "acs5", cache = TRUE)
```

#### Pulling in spatial data 

tidycensus can pull in the associated spatial data while simultaneously pulling in the ACS/Census data.  

This is specified by adding `geometry = TRUE`   

Note that the default geographic coordinate system is NAD 1983, which you may want to change.  

Let's look at the median household income from the 2016-2020 ACS for Hennepin county again, but this time pull in the simple feature geometry as well.  

```{r, results = HIDE}
hennepin <- get_acs(
  state = "MN",
  county = "Hennepin",
  geography = "tract",
  variables = c(median_income = "B19013_001"), # You can rename the variable here
  geometry = TRUE, #Adding in the spatial component
  year = 2020
)

head(hennepin)
```

#### Making maps!

Since tidycensus uses tidy formatting, the objects can be easily visualized with ggplot2

Let's make a map of Hennepin county with median income, from data pulled above:
```{r}
hennepin %>%
  ggplot(aes(fill = estimate)) + 
  geom_sf(color = "black") + 
  scale_fill_viridis_c(option = "magma") 
```

### Using tidyverse tools to manipulate Census/ACS data

#### Using summary variables to do comparisons

##### Relativizing your data

Much of the ACS and Census data are counts, which are can't really be readily compared. Most people are interested in how these data for a given geography compare *relative* to another. 

As a solution, you can normalize counts by the total population size, which are often found in the same table. This can be done using `summary_var` as an additional argument. 

Let's look at race for Hennepin county. Use `view()` to browse table `B03002` in the ACS data from 2016-2020. From here you can see that the variable `B03002_001' gives us total population.  Using this information, we can compare the population of tracts in Hennepin county by race & Hispanic origin with the total population size.  

```{r}

# Renaming the ACS variables from Table B03002 into more useful names 
race_vars <- c(
  White = "B03002_003",
  Black = "B03002_004",
  Native = "B03002_005",
  Asian = "B03002_006",
  HIPI = "B03002_007",
  Hispanic = "B03002_012"
)


hennepin_race <- get_acs(
  state = "MN",
  county = "Hennepin",
  geography = "tract",
  variables = race_vars,
  summary_var = "B03002_001", # Including the total population as a denominator
  geometry = TRUE, #Adding in the spatial component
  year = 2020
)

```

```{r, eval = FALSE, echo = FALSE}
hennepin_race %>%
  ggplot(aes(fill = summary_est), group = variable) + 
  geom_sf(color = "black") + 
  scale_fill_viridis_c(option = "magma") 

```


##### Calculating percentages

Now that are data are relativized, we can used `mutate()` to calculate a percentage column. This column would correspond to the percentage of each tract's population that is represented by each racial/ethnic group in 2016-2020.

```{r}

hennepin_race_percent <- hennepin_race %>%
                         mutate(percent = 100 * (estimate / summary_est)) %>%
                         select(NAME, variable, percent) # Only keep columns of interest and dropping the rest
head(hennepin_race_percent)

```

